Learning and Reputation under Monopoly and Duopoly: \\A Competing Bandits Approach

### 2018-08-28 paper/competing_bandits.tex
edits on intro

\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\usepackage{graphicx}
\frenchspacing

@@

% a very useful package for edits and comments, from David Kempe (USC)
\usepackage{color-edits}
%\usepackage[suppress]{color-edits}  % use this to suppress the package
\addauthor{as}{red}    % as for Alex
\addauthor{ga}{blue}  % ga for Guy
\addauthor{sw}{green} % sw for Steven
% e.g. for Alex, provides \asedit{}, \ascomment{} and \asdelete{}.

@@

Platforms routinely deploy randomized controlled trials (A/B tests), and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known framework for exploration and making decisions under uncertainty. While deploying ``better" learning algorithms for exploration would (deployment/maintenance costs aside) improve performance, this is not necessarily beneficial under competition. In particular, excessive experimentation may hurt platform's reputation and decrease market share in the near term. This would leave the learning algorithms with less users to learn from, which may further degrade platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth.

We ask whether competition incentivizes adoption of "better" algorithms for exploration. We investigate this issue via extensive numerical experiments in a stylized duopoly model. In our model, two firms compete for users, and simultaneously learn from them. Each firm commits to a multi-armed bandit algorithm, and \emph{explores} according to this algorithm. Users select between the two firms based on the current reputation score: rewards from the firm's algorithm, averaged over a recent time window. Each firms' objective is the market share.

We consider varying levels of competition, from monopoly to \emph{temporary monopoly} (a duopoly with an early entrant), to a true duopoly in which both firms start at the same time. We find that a ``greedy algorithm" that does not explicitly explore is most beneficial under duopoly. This algorithm also prevails under monopoly simply because it tends to be easier to deploy. Whereas a temporary monopoly incentives more advanced exploration algorithms that perform better in the long run. These findings can be framed in terms of an ``inverted-U relationship" between competition and innovation.

@@

>>>>

\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\usepackage{graphicx}
\usepackage{tikz} % SW: Added for plots
\frenchspacing

@@

% a very useful package for edits and comments, from David Kempe (USC)
\usepackage{color-edits}
%\usepackage[suppress]{color-edits}  % use this to suppress the package
\addauthor{as}{red}    % as for Alex
\addauthor{ga}{blue}  % ga for Guy
\addauthor{sw}{orange} % sw for Steven
% e.g. for Alex, provides \asedit{}, \ascomment{} and \asdelete{}.

@@

Platforms routinely deploy randomized controlled trials (A/B tests), and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known framework for exploration and making decisions under uncertainty. While deploying ``better" learning algorithms for exploration would improve performance, this is not necessarily beneficial under competition \swedit{(putting deployment/maintenance costs aside)}. In particular, excessive experimentation may hurt platform's reputation and decrease market share in the near term. This would leave the learning algorithms with less users to learn from, which may further degrade platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth.

We ask whether competition incentivizes adoption of "better" algorithms for exploration. We investigate this issue via extensive numerical experiments in a stylized duopoly model. In our model, two firms compete for users, and simultaneously learn from them. Each firm commits to a multi-armed bandit algorithm, and \emph{explores} according to this algorithm. Users select between the two firms based on the current reputation score: rewards from the firm's algorithm, averaged over a recent time window. \swedit{The objective of both firms is to maximize the market share ---the fraction of users coming to them.}

We consider varying levels of competition, from monopoly to \emph{temporary monopoly} (a duopoly with an early entrant), to a true duopoly in which both firms start at the same time. We find that a ``greedy algorithm" that does not explicitly explore is most beneficial under duopoly. This algorithm also prevails under monopoly simply because it tends to be easier to deploy. Whereas a temporary monopoly incentives more advanced exploration algorithms that perform better in the long run. These findings can be framed in terms of an ``inverted-U relationship" between competition and innovation (see Figure~\ref{fig:inverted-U}).

\tikzstyle{level 1}=[level distance=3.5cm, sibling distance=4.0cm]
\tikzstyle{level 2}=[level distance=3.5cm, sibling distance=2cm]
\tikzstyle{below} = [align=center]

\begin{figure}
\begin{center}
\begin{tikzpicture}
      \draw[->] (-.5,0) -- (6,0) node[right] {};
      \draw[->] (0,-.5) -- (0,2) node[above] {Better algorithms};
      \draw[scale=0.6,domain=0.7:8.3,smooth,variable=\x,blue, line width=0.3mm] plot ({\x},{3.4 - 0.2 * (\x - 4.5)^2});
     \node[below] at (0.7, -0.15) {\footnotesize monopoly};
     \node[below] at (3, 0) {\footnotesize temporary \\ monopoly};
     \node[below] at (5, -0.15) {\footnotesize duopoly};
 \end{tikzpicture}
 \caption{Inverted-U relationship between different regimes of
   competition and the incentivized exploration algorithms.}
\label{fig:inverted-U}
\end{center}
\end{figure}

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
